---
title: "Modelling Biological Systems Exam 2016"
author: "Sergii Gladchuk"
date: "December 29, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2. Optimal fishing of an unstructured population.

Consider a fish population that grows according to

$$\frac{dn}{dt}=rn\left(1-\left(\frac{n}{K}\right)^2\right),$$

where $r$ and $K$ are positive constants.

## a) What is the (non-trivial) equilibrium population size?

Solve $f(n^*) = 0$

Trivial is $n = 0$

Non-trivial:
$$1 - \left(\frac{n}{K}\right)^2 = 0$$
$$n = K$$

since K is positive constant

## b) Show that it is a stable equilibrium.

An equilibrium is stable if $f'(n^*) < 0$

$$f'(n) = f'(rn)  \left(1 -\left(\frac{n}{K}\right)^2\right) + rn  f'\left(1 -\left(\frac{n}{K}\right)^2\right) = r - \frac{rn^2}{K^2} - \frac{2rn^2}{K^2} = r - \frac{3rn^2}{K^2}$$

$$f'(n^*) = r - 3r = -2r < 0$$
since r is positive constant.

The same shows Growth Function graph based on our model - the slope is negative at $n = K$ point:

```{r}
library(deSolve)

#define the drowth function

populationGrowth <- function(t, n, P) {
  dndt <- P$r0 * n * (1 - n**2/P$K**2)
  list(dndt)
  
}

# set the vector of time-points for the output
timevec <- seq(0, 20, by=0.1)

# list of parameters

P <- list(r0 = 1.3, K = 100)

n0 <- 2 #initial population size

# call the ode function
n <- 1:110
out <- ode(y = n0, func = populationGrowth, times = timevec, parms = P)
dndt <- unlist(populationGrowth(1, n, P))
par(mfrow = c(1,2))
plot(out[,1], out[,2], type='l', main = 'Population growth', 
     xlab='Time', ylab='Population size')
plot(n, dndt, type='l', main='Growth Function', xlab='n', ylab='dn/dt',col='red')
abline(0,0)

```
\pagebreak

Now assume the population is harvested, such that a proportion h is harvested per
time unit.

## c) Add the harvesting to the population dynamic model!

$-hn$ should be added to starting model

$$\frac{dn}{dt}=rn\left(1-\left(\frac{n}{K}\right)^2\right) - hn$$

## d) Where is the new equilibrium population size?

Solve $f(n^*) = 0$

$$n\left(r\left(1 - \left(\frac{n}{K}\right)^2\right) - h\right) = 0$$

Trivial is $n = 0$

Non-trivial:
$$r\left(1 - \left(\frac{n}{K}\right)^2\right) - h = 0$$

$$r - \frac{rn^2}{K^2} - h = 0$$

$$\frac{r}{K^2}  n^2 = r - h$$

$$n^2 = K^2  \left(1 - \frac{h}{r}\right)$$

$$n = K \sqrt{1 - \frac{h}{r}}$$

Confirmation by running in R

```{r}
library(deSolve)
#define the drowth function with fishing
populationGrowth <- function(t, n, P) {
  dndt <- P$r0 * n * (1 - n**2/P$K**2) - n*P$h
  list(dndt)
}
# set the vector of time-points for the output
timevec <- seq(0, 20, by=0.1)
# list of parameters
P <- list(r0 = 1.5, K = 100, h = 0.5)
n0 <- 50 #initial population size
# call the ode function
out <- ode(y = n0, func = populationGrowth, times = timevec, parms = P)
n <- 1:110
dndt <- unlist(populationGrowth(1, n, P))
par(mfrow = c(1,2))
plot(out[,1], out[,2], type='l', main = 'Population growth', xlab='Time', ylab='Population size')
plot(n, dndt, type='l', main='Growth Function', xlab='n', ylab='dn/dt',col='red')
abline(0,0)

```

There is stable equilibrium point (lower than K) with negative slope.

## e) At what harvest rate does the population go extinct?

If $h > r \Rightarrow n < 0$

This is confirmed by running same model with $h > r$

```{r}
library(deSolve)
#define the drowth function with fishing
populationGrowth <- function(t, n, P) {
  dndt <- P$r0 * n * (1 - n**2/P$K**2) - n*P$h
  list(dndt)
}
# set the vector of time-points for the output
timevec <- seq(0, 20, by=0.1)
# list of parameters
P <- list(r0 = 1.5, K = 100, h = 1.7)
n0 <- 50 #initial population size
# call the ode function
out <- ode(y = n0, func = populationGrowth, times = timevec, parms = P)
n <- 1:110
dndt <- unlist(populationGrowth(1, n, P))
par(mfrow = c(1,2))
plot(out[,1], out[,2], type='l', main = 'Population growth', xlab='Time', ylab='Population size')
plot(n, dndt, type='l', main='Growth Function', xlab='n', ylab='dn/dt',col='red')
abline(0,0)

```

There is no second stable equilibrium point on Growth Function graph

\pagebreak

## f) At what harvest rate is the yield, i.e. the total number of harvested individuals, maximized?

Yield ($hn$) at equilibrium:

$$hK\sqrt{1 - \frac{h}{r}}$$

Maximized yield can be found at 0 point for derivative of above function $f'(h^*) = 0$

$$f'(h) = f'(Kh) \sqrt{1 - \frac{h}{r}} + kh f'\left(\sqrt{1 - \frac{h}{r}}\right) = K \sqrt{1 - \frac{h}{r}} + Kh\frac{1}{2}\left(1 - \frac{h}{r}\right)^{-\frac{1}{2}} \left(-\frac{1}{r}\right) = $$

$$= K\sqrt{1 - \frac{h}{r}} - \frac{Kh}{2r\sqrt{1 - \frac{h}{r}}} = \frac{2rK\left(1-\frac{h}{r}\right) - Kh}{2r\sqrt{1 - \frac{h}{r}}} = K \frac{2r - 3h}{2r\sqrt{1 - \frac{h}{r}}}$$

So for $f'(h^*) = 0$ either $K = 0$ or

$$\frac{2r - 3h}{2r\sqrt{1 - \frac{h}{r}}} = 0$$
$$2r - 3h = 0$$
$$h = \frac{2}{3}r$$

